Under-ice acoustic navigation using real-time model-aided range estimation

### 2021-08-26 doc/main.tex
update title per Henrik's email
minor/general edits and suggestions

\title[JASA/draft]{Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation}
\author{EeShan C. Bhatt}
\email{ebhatt@whoi.edu}
\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Oscar Viquez}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Henrik Schmidt}
\email{henrik@mit.edu}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}

@@

Autonomous underwater vehicles (AUVs) are an increasingly capable platform to explore and sample the ocean, particularly for remote and/or dangerous environments.
Lightweight, mobile, and inexpensive compared to crewed- and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.

@@

Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}. 
The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020} and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.

@@

The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more incorrect due to increased multipath in the upper duct of the Beaufort Lens.

The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
The challenges anticipated during ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.

@@

The acoustic communication dataset accrued for ICEX20 relied on the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, Arctic ice, and ocean acoustics; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software and hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
Both simulation capabilities are integral to mission success.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
	\caption{A cartoon overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data and tracking information between AUV and a human decision maker at Topside.}
	\label{fig:icnnOverview}
\end{figure}

@@

The design enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
This adaptivity was not used during the modem experiment but was instead toggled manually.
Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were transmitting simultaneously recorded events as modems, not as vehicles.

@@

During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a CTD, i.e. an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B. Howard and LT CDR D. Goodwin).
This order corresponds to increasing ducted conditions of the Beaufort Lens.

@@

When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
The group velocity estimate is found using the extrapolated navigation for range, $\hat{r}$, and an impulse response estimate received in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.

The initial call to BELLHOP is over a sparse local grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
\begin{equation}
u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
\end{equation}
and the associated weighted standard deviation,
\begin{equation}
\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
\end{equation}
If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.

@@

It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
The analysis here therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.

@@

\begin{figure}[h!]
\includegraphics[width=\columnwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis. The left panel is for a receiver depth of 30 m; the right panel for 90 m. The sound speed source is indicated by color. The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored orange dots showing the naive, data-driven group velocity calculation.}
\label{fig:gvel30}
\end{figure}

@@

The goal of the group velocity estimation is to converge towards the naive group velocity estimation, i.e. the GPS-derived range divided by the recorded travel time.
For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated time arrival packet.

@@

Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
The weighted SSP range error generally has the smallest and most zero-centered footprint.
The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts propagating through the steep sound speed gradients of the lens and through the shadow zone. 
All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.

@@

The large error is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small, at 0.025\%. 
This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.

@@

% operational
There are two metrics that demonstrate the operational limitations of GPS tracking technologies\textemdash fix rate and positional dilution of precision (PDOP) \citep{swanlund_gps_2016}.
Fix rate looks at the ratio of successful fixes to overall attempts in acquiring a location.
In animal tracking studies in polar latitudes, fix rates have significantly improved in the last twenty years, from as low as 26\% in 1996 \citep{moen_effects_1996} to 87.2\% in 2018 \citep{jung_accuracy_2018}.
The latter study showed a precision of 4.3 $\pm$ 4.0 m and an accuracy of 6.0 $\pm$ 4.7 m.
While significant metric to evaluate GPS robustness, accuracy and precision, not the fix rate, explains the severe banding observed; fix rate bias, where various aspects of a landscape can induce bias by impeding the GPS receiver from acquiring its location, is not relevant given the static nature and small time window of ICEX20.

@@

These are just two subsets of the physical links that cover all four GPS modem buoys, except for Camp.
The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.

@@

There are many avenues that this approach can be further refined and tested for field operations.
Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range.

>>>>

\title[JASA/draft]{Under-ice acoustic navigation using real-time model-aided range estimation}
\author{EeShan C. Bhatt}
\email{ebhatt@whoi.edu}
\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Oscar Viquez}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Henrik Schmidt}
\email{henrik@mit.edu}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}

@@

Autonomous underwater vehicles (AUVs) are an increasingly capable platform to explore and sample the ocean, particularly for remote and/or dangerous environments.
Lightweight, mobile, and inexpensive compared to crewed and ship-based systems, these vehicles can enable a safer, data-forward style of field oceanography \citep{bellingham_robotics_2007,petillot_underwater_2019}.

@@

Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}. 
The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.

@@

The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more inaccurate due to increased multipath in the upper duct of the Beaufort Lens.

The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
The challenges anticipated for ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.

@@

The acoustic communication dataset accrued for ICEX20 relied on the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, including Arctic ice drift and ocean acoustic propagation; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software-only configuration and a hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
Both simulation capabilities are integral to mission success.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
	\caption{A schematic overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data-transfer and tracking between AUV and a human decision maker at Topside.}
	\label{fig:icnnOverview}
\end{figure}

@@

The design enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
This adaptivity was not used during the modem experiment but was instead toggled manually.
Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were active during the experiment recorded events as tracking buoys, not as vehicles.

@@

During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a local CTD cast, i.e. an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B. Howard and LT CDR D. Goodwin).
This order corresponds to increasing ducted conditions of the Beaufort Lens.

@@

When the topside tracking framework receives a modem message, with a time delay, $\Delta t$, from one of the range modems, it will request a new estimate of the group velocity and its associated uncertainty.
The group velocity estimate is found using the vehicle's reported depth and the extrapolated navigation solution for range, $\hat{r}$, as inputs for the ray tracing program. The latter returns an impulse response estimate in the form of ray travel times $dt_{j}$ and amplitudes $a_{j}$ for that range and depth.

The initial call to BELLHOP is over a sparse local grid, from which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
\begin{equation}
u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
\end{equation}
and the associated weighted standard deviation,
\begin{equation}
\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
\end{equation}
If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.

@@

It is important to note that the value for the extrapolated range, $\hat{r}$, is only tracked when the modem runs the vehicle behavior; thus we replace $\hat{r}$ with the GPS-tracked range for all modem events.
Because $\hat{r}$ converges to the correct solution, a comparison of $\hat{r}$ with the GPS-tracked range shows a normal, zero-centered distribution within the bounds of GPS drift.
The present analysis therefore seeds realistic but ``omniscient'' knowledge of the extrapolated range and leverages the post-processing pipeline to more thoroughly evaluate the acoustic range estimate for all modem events, with three relevant sound speed sources, and both group velocity criterion.
Accordingly, the results in this section evaluate the utility of the algorithms and sound speed sources, divorced from their role in the ICNN while maintaining real-time relevance.

@@

\begin{figure}[h!]
\includegraphics[width=\columnwidth]{figs/SLIDES-gvel-comparison-wdata-30.pdf}
\caption{A comparison of group velocity predictions for all beacon to beacon events in post-processing with a source depth of 30 m, with group velocity on the y-axis and recorded travel time on the x-axis. The left panel is for a receiver depth of 30 m; the right panel for 90 m. The sound speed source is indicated by color. The minimal and nearest bounce criterion are distinguished by different marker shapes, compared to the separately colored red dots showing the naive, data-driven group velocity calculation.}
\label{fig:gvel30}
\end{figure}

@@

The goal of the group velocity estimation is to converge towards the naive group velocity estimation, i.e. the GPS-derived range divided by the recorded travel time.
For a 30 m receiver depth, the NBC shows more overlap with data-derived values as it classifies multipath more correctly.
For a 90 m receiver depth, the overlap is less accurate due to computational constraints of a limited fan of rays entering the shadow zone rendering a less reliable simulated times of arrival packet.

@@

Fig. \ref{fig:rangeError} shows the directional range error footprints for all three sound speed inputs with respect to OWTT, separated by source and receiver depth configurations.
The weighted SSP range error generally has the smallest and most zero-centered footprint.
The one case it does not is for the source-receiver pairings between 30 and 90 m in depth.
The increased error for these reciprocal transmission paths is most likely driven by the computational artifacts encountered when propagating through the steep sound speed gradients of the lens and through the shadow zone.
All other source depth pairings are significantly improved using the chosen weights compared to HYCOM or the baseline.

@@

The large error in this instance is driven by the MBC unexpectedly defaulting to a bottom bounce with a much greater OWTT.
The NBC classifies the multipath as 4 bounces, reducing the range error from greater than a kilometer to less than a meter.
While there is no actual way of knowing if this is the correct multipath structure, the range error is remarkably small, at 0.025\%. 
This pattern of not choosing the minimal observed bounce structure is consistent across all SSPs; the baseline goes from 1 to 3 bounces and HYCOM goes from 0 to 2 bounces.
Notably, the baseline and HYCOM range errors are never egregiously large, but are nonetheless improved with the NBC algorithm.
Thus, for acoustically complex environments, the NBC has a disproportionately positive impact as the estimated SSP approaches the desired SSP.

@@

% operational
There are two metrics that demonstrate the operational limitations of GPS tracking technologies\textemdash fix rate and positional dilution of precision (PDOP) \citep{swanlund_gps_2016}.
Fix rate looks at the ratio of successful fixes to overall attempts in acquiring a location.
In animal tracking studies in polar latitudes, fix rates have significantly improved in the last twenty years, from as low as 26\% in 1996 \citep{moen_effects_1996} to 87.2\% in 2018 \citep{jung_accuracy_2018}.
The latter study showed a precision of 4.3 $\pm$ 4.0 m and an accuracy of 6.0 $\pm$ 4.7 m.
While fix rate is a significant metric to evaluate GPS robustness, it is accuracy and precision, and not the fix rate, that explain the severe banding observed; fix rate bias, where various aspects of a landscape can induce bias by impeding the GPS receiver from acquiring its location, is not relevant given the static nature and small time window of ICEX20.

@@

These are just two subsets of the physical links that cover all four GPS modem buoys.
The GPS at camp is the least accurate due to the human activity and infrastructure occluding the physical puck.

@@

There are many avenues through which this approach can be further refined and tested for field operations.
Amongst them is defining the uncertainty grid for BELLHOP via stochastic or data-driven measures such as the distance traveled by the AUV between ICNN updates or the magnitude of the position corrections by the ICNN.
Another is to entirely remove the seeded range and instead rely on the submerged asset's depth and recorded OWTT to find high probability fields in range.
